A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-20 model.tex
added loren's instability

We suggest a useful way to understand dust-drag instabilities in
protoplanetary disks is to re-cast the problem into that of
thermodynamics. Then such instabilities appear in the equations in the
same way as pulsational instabities in stars, which has already been
well studied.

>>>>

We suggest understanding dusty instabilities in
protoplanetary disks by re-casting the problem into that of
thermodynamics. Then dust-drag instabilities appear in the equations in the 
same way as pulsational instabities in stars, which has already been 
well studied. Whether or not the dusty disk is actually unstable
depends on details of the dust-drag or `cooling' function
$\mathcal{C}$. However, we might generally expect the system to
support a range of modes. On physical grounds, unstable modes 
\emph{must} be so by causing a phase lag between pressure and total
density.

 %Whether the phase
%difference is positive or negative depends on the cooling function
%$\mathcal{C}$, but we may expect a system to generally have both
%modes, and that with a phase lag are unstable.

### 2016-08-20 paper.tex
added loren's instability

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  considered as an effective 
  heating/cooling. We exploit this equivalence 
  to 
%\begin{inparaenum}[1)]
%\item
%  identify the entropy of isothermal dusty gas; and 
%\item 
  generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%\end{inparaenum}
%This includes the axisymmetric stability of strictly isothermal dusty
%gas, allowing us to constrain the sharpness of 
%dust edges that can persist in protoplanetary disks. 
We show that strictly isothermal, non-self-gravitating, axisymmetric
disks perfectly coupled to dust are generally stable, however thin the
dust layer is.  %except possibily in the presence of sharp 
%edges in radial dependence of the dust-to-gas ratio. 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss thermodynamical interpretations of 
dust-gas drag dynamics, such as the streaming instability in
protoplanetary disks as a manifestation of pulsational instabilities
in stars.  
\end{abstract}

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  considered as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show in the limit of perfect coupling, that  dust-settling cannot lead to axisymmetric instabilities, however thin the dust layer become.     
%
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss thermodynamical interpretations of 
dust-gas drag dynamics, such as the streaming instability in
protoplanetary disks as a manifestation of pulsational instabilities
in stars.  
\end{abstract}

### 2016-08-20 summary.tex
added loren's instability

\subsection{Understanding dust-drag
  instabilities}\label{dust_as_thermo}

>>>>

\subsection{Other dust-drag
  instabilities}\label{dust_as_thermo}

As described in \S\ref{dust_work}, 
the physical cause of dust-drag instabilities is a phase lag between
the pressure and density response in a dusty gas\footnote{
In fact, we expect a phase difference between pressure and the total
density  whenever $\tstop\neq0$; not just $\tstop\OmK\ll 1$ as 
considered in this work. This is simply because of the finite
relaxation time required for dust to respond to gas, and vice versa. 
An effective energy equation in the form of Eq. \ref{eff_energy} can
still be obtained for general $\tstop$. The
heat flux $\bm{F}$ in the cooling function $\mathcal{C}$ is determined
through solution to the relative velocity $\Delta \bm{v} =
\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$ \citep{youdin05a,laibe14}.  We
thus expect the `pressure-density lag' interpretation of dust-drag 
instabilities to be generally applicable.}. 
This may be used to re-interpret the streaming instability or to
understand other dusty instabilities. For example,
\cite{loren15,loren16} have recently discovered a new axisymmetric 
instability in their two-fluid numerical simulations of dusty
protoplanetary disks. They find it leads to the formation of
`toroidal vortices'.

%give numerical example 
We ran our one-fluid stability code to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by \citeauthor{loren15}, as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity $Z=0.01$ and dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The disk parameters are $p=-1.5,\, q=-1$, and  
$\hgas=0.05$. We set $k = \pi/H_\mathrm{d}$.

Fig. \ref{result2d_loren} shows an unstable mode found for this setup,
with  
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. This is not a VSI mode, as we
find it persists when $q=0$. We estimate a much smaller growth
rate for the streaming instability, of 
$O(10^{-3}\hgas\OmK)$, for the above parameters. We thus also rule out
SI, as did \citeauthor{loren15}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result2d_loren.ps}
  \caption{Possible eigenmode of the dusty instability discovered by 
    \cite{loren15} in their numerical simulations. The growth rate and 
    radial spacing of the `toroidal vortices' are similar to that
    reported by \citeauthor{loren15}. The instability is driven by
    dust-gas drag. The color scale shows the
    perturbed  
    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$.
    \label{result2d_loren}
  }
\end{figure}

This preliminary calculation only serves to highlight the possibility
of new axisymmetric dust-drag instabilities in stratified disks. One 
issue is that the base state for the above calculation is not a true
equilibrium (although \citeauthor {loren15} find the dust settling
time is much longer than the instability growth time). We will study
this instability in more detail with a proper basic
state in a follow up work.
